---
title: "Data Manipulation with 'dplyr'"
output: github_document
date: "2022-09-22"
---

Once you've imported data, you're going to need to do some cleaning up.

```{r}
knitr::opts_chunk$set(
  collapse=TRUE,
  fig.width = 6
)
library(tidyverse)
```

```{r}
options(tibble.print_min = 3)

litters_data = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data = janitor::clean_names(litters_data)
view

pups_data = read_csv("./data/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)
```

## Select

You can specify the columns you want to keep by naming all of them:
```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
```

You can specify the specify a range of columns to keep:
```{r}
select(litters_data, group:gd_of_birth)
```


You can also specify columns you’d like to remove:
```{r}
select(litters_data, -pups_survive)
```

You can rename variables as part of this process:
```{r}
select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```


If all you want to do is rename something, you can use rename instead of select. This will rename the variables you care about, and keep everything else:
```{r}
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

There are some handy helper functions for select; read about all of them using ?select_helpers. I use starts_with(), ends_with(), and contains() often, especially when there variables are named with suffixes or other standard patterns:
```{r}
select(litters_data, starts_with("gd"))
```

I also frequently use is everything(), which is handy for reorganizing columns without discarding anything:
```{r}
select(litters_data, litter_number, pups_survive, everything())
```

relocate does a similar thing (and is sort of like rename in that it’s handy but not critical):
```{r}
relocate(litters_data, litter_number, pups_survive)
```

## Filter

You will often filter using comparison operators (>, >=, <, <=, ==, and !=). You may also use %in% to detect if values appear in a set, and is.na() to find missing values. The results of comparisons are logical – the statement is TRUE or FALSE depending on the values you compare – and can be combined with other comparisons using the logical operators & and |, or negated using !.

Some ways you might filter the litters data are:

* gd_of_birth == 20
* pups_born_alive >= 2
* pups_survive != 4
* !(pups_survive == 4)
* group %in% c("Con7", "Con8")
* group == "Con7" & gd_of_birth == 20

```{r}
filter(litters_data, gd_of_birth == 20)
```

```{r}
filter(litters_data, group == "Con7" & gd_of_birth == 20)
```

A very common filtering step requires you to omit missing observations. You can do this with filter, but I recommend using drop_na from the tidyr package:

* drop_na(litters_data) will remove any row with a missing value
* drop_na(litters_data, wt_increase) will remove rows for which wt_increase is missing.



## Mutate

The example below creates a new variable measuring the difference between gd18_weight and gd0_weight and modifies the existing group variable.
```{r}
mutate(litters_data,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)
```

A few things in this example are worth noting:

Your new variables can be functions of old variables
New variables appear at the end of the dataset in the order that they are created
You can overwrite old variables
You can create a new variable and immediately refer to (or change) it

## Arrange

You can arrange the rows in your data according to the values in one or more columns:
```{r}
head(arrange(litters_data, group, pups_born_alive), 10)
```

## Pipes

For example, suppose you want to load the data, clean the column names, remove pups_survive, and create wt_gain. There are a couple of options for this kind of multi-step data manipulation:

First option: define intermediate datasets (or overwrite data at each stage)
```{r}
litters_data_raw = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data_clean_names = janitor::clean_names(litters_data_raw)
litters_data_selected_cols = select(litters_data_clean_names, -pups_survive)
litters_data_with_vars = 
  mutate(
    litters_data_selected_cols, 
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group))
litters_data_with_vars_without_missing = 
  drop_na(litters_data_with_vars, wt_gain)
litters_data_with_vars_without_missing
```

Second option: nest function calls
```{r}
litters_data_clean = 
  drop_na(
    mutate(
      select(
        janitor::clean_names(
          read_csv("./data/FAS_litters.csv", col_types = "ccddiiii")
          ), 
      -pups_survive
      ),
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
    ),
  wt_gain
  )

litters_data_clean
```

Best option: Piping!
```{r}
litters_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(wt_gain)

litters_data
```

All three approaches result in the same dataset, but the piped commands are by far the most straightforward. The easiest way to read %>% is “then”; the keyboard shortcuts are Ctrl + Shift + M (Windows) and Cmd + Shift + M (Mac).

The functions in dplyr (and much of the tidyverse) are designed to work smoothly with the pipe operator. By default, the pipe will take the result of one function call and use that as the first argument of the next function call; by design, functions in dplyr will take a tibble as an input and return a tibble as a result. As a consequence, functions in dplyr are easy to connect in a data cleaning chain. You can make this more explicit by using . as a placeholder for the result of the preceding call:

```{r}
litters_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names(dat = .) %>%
  select(.data = ., -pups_survive) %>%
  mutate(.data = .,
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(data = ., wt_gain)
```

In this example, the dataset argument is called dat in janitor::clean_names, .data in the dplyr functions, and data in drop_na – which is definitely confusing. In the majority of cases (and everywhere in the tidyverse) you’ll elide the first argument and be happy with life, but there are some cases where the placeholder is necessary. For example, to regress wt_gain on pups_born_alive, you might use:
```{r}
litters_data %>%
  lm(wt_gain ~ pups_born_alive, data = .) %>%
  broom::tidy()
```


